---
title: "AntTraits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AntTraits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Load the packages and get the ant traits data from gllvm

```{r setup, message=FALSE}
library(HierOrd)
library(INLA)
library(gllvm)
data("antTraits")
```

The data has abundances of `r ncol(antTraits$abund)` species of ant on `r nrow(antTraits$abund)` sites. In addition there are `r ncol(antTraits$env)` environmenral variabls and `r ncol(antTraits$traits)` traits. We scale the 

```{r}
ScaleNumeric <- function(df) {
  IsNum <- which(unlist(lapply(df, is.numeric), use.names = FALSE))
   x2 <- data.frame(df)
  x2[,IsNum] <- apply(df[,IsNum], 2, scale)
  x2
}
 
# in FitHierOrd(), these lines need to have covariates & factors passed to them correctly:
# Might have to re-write CreateDataframe() to change output
# row.formula <- MakeFormula(nlv=nLVs, linnames=data$Names$LVCovs$Row, epsname="Row",
#                              factprior = RowCovHyper, epsprior = EpsHyper)
#   col.formula <- MakeFormula(nlv=nLVs, linnames=data$Names$LVCovs$Col, epsname="Col",
#                              factprior = RowCovHyper, epsprior = EpsHyper)

antTraits$envSc <- ScaleNumeric(antTraits$env)
antTraits$traitsSc <- ScaleNumeric(antTraits$traits)
Res <-FitHierOrd(Y=antTraits$abund, rowdata=antTraits$envSc, 
                 coldata=antTraits$traitsSc, nLVs = 1, 
                  family = "Poisson", NBurnin = 10, NIter = 100, NChains = 4, 
#                  family = "Poisson", NBurnin = 1, NIter = 2, NChains = 2, 
                  INLAobj=TRUE)

RowMergeRes1 <- inla.merge(Res$rowinla[[1]])
RowMergeRes2 <- inla.merge(Res$rowinla[[1]])
RowMergeRes3 <- inla.merge(Res$rowinla[[1]])
RowMergeRes4 <- inla.merge(Res$rowinla[[1]])
RowMergeRes <- inla.merge(list(RowMergeRes1, RowMergeRes2, RowMergeRes3, RowMergeRes4))

ColMergeRes1 <- inla.merge(Res$colinla[[1]])
ColMergeRes2 <- inla.merge(Res$colinla[[2]])
ColMergeRes3 <- inla.merge(Res$colinla[[3]])
ColMergeRes4 <- inla.merge(Res$colinla[[4]])
ColMergeRes <- inla.merge(list(ColMergeRes1, ColMergeRes2, ColMergeRes3, ColMergeRes4))
rm(RowMergeRes1, RowMergeRes2, RowMergeRes3, RowMergeRes4, 
   ColMergeRes1, ColMergeRes2, ColMergeRes3, ColMergeRes4)

save(Res, RowMergeRes, ColMergeRes, file="../inst/AntsRes.RData")

```


<details> <summary>The mixing looks OK</summary>

Rows first:

```{r, fig.height=8}
plot(Res$col[,6:8])

```

Columns

```{r, fig.height=8}
plot(Res$col[,6:8])

```

</details>

Now we can plot site and species scores. Sorry for the species names dropping off.


```{r PlotTrueRow, fig.width=8, fig.height=8}
RowMean <- summary(Res$row)$statistics[,"Mean"]
RowCI <- summary(Res$row)$quantiles

ColMean <- summary(Res$col)$statistics[,"Mean"]
ColCI <- summary(Res$col)$quantiles
SpNames <- gsub("  ", ". ", gsub("\\.", " ", colnames(antTraits$abund)))
par(mfrow=c(1,2), mar=c(2,2,3,1), oma=c(2,8,0,0))
plot(ColMean, 1:length(ColMean), xlim=range(ColCI), main="Column", yaxt="n", ann=FALSE)
axis(2, SpNames, at=1:length(ColMean), las=1)
segments(ColCI[,"2.5%"], 1:length(ColMean), ColCI[,"97.5%"], 1:length(ColMean))
# abline(0,1, col=2); abline(0,-1, col=2)

plot(RowMean, 1:length(RowMean), xlim=range(RowCI), main="Row", yaxt="n", ann=FALSE)
axis(2, rownames(antTraits$abund), las=1)
segments(RowCI[,"2.5%"], 1:length(RowMean), RowCI[,"97.5%"], 1:length(RowMean))
# abline(0,1, col=2); abline(0,-1, col=2)



```

Now we can look at the covariate effects. We first merge the INLA runs, to marginalise over the row/column effects that were simulated.

```{r RowCovResults, warning=FALSE}
# RowMergeRes <- inla.merge(do.call(c, Res$rowinla))
RowHPD.l <- lapply(RowMergeRes$marginals.fixed, function(marg, p) inla.hpdmarginal(p, marg), p=0.95)
RowHPD <- cbind(
  At = 1:length(RowMergeRes$marginals.fixed),
  mode = unlist(lapply(RowMergeRes$marginals.fixed, inla.mmarginal)),
  low = unlist(lapply(RowHPD.l, function(x) x[1,"low"])),
  high = unlist(lapply(RowHPD.l, function(x) x[1,"high"]))
)
dimnames(RowHPD)[[1]] <- gsub("\\.", " ", gsub(".LV.*", "", dimnames(RowHPD)[[1]]))
```

Then we plot:

```{r PlotRowRes}
par(mar=c(4.1,9,1,1))
plot(RowHPD[-1, "mode"], RowHPD[-1, "At"], xlim=range(RowHPD[-1, c("low", "high")]),
     ann=FALSE, axes=FALSE)
segments(RowHPD[-1, "low"], RowHPD[-1, "At"], RowHPD[-1, "high"], RowHPD[-1, "At"])
abline(v=0)
axis(1)
axis(2, dimnames(RowHPD)[[1]][-1], at=RowHPD[-1, "At"], las=1)

```

Which suggests that shrub cover has an effect: more cover, more of some stuff.

Now traits. Note that Pilosity and Polymorphism should be factors.

```{r ColCovResults, warning=FALSE}
# ColMergeRes <- inla.merge(do.call(c, Res$colinla))
#Fix <- summary(ColMergeRes)$fixed
ColFixedHPD.l <- lapply(ColMergeRes$marginals.fixed, function(marg, p) inla.hpdmarginal(p, marg), p=0.95)
ColRandHPD.l <- lapply(ColMergeRes$marginals.random[1:2], function(lst) 
  lapply(lst, function(marg, p) inla.hpdmarginal(p, marg), p=0.95))

ColRandHPD.l2 <- do.call(c, unlist(ColRandHPD.l, recursive=FALSE))

ColRandmode <- unlist(lapply(ColMergeRes$marginals.random[1:2], function(lst) 
  lapply(lst, inla.mmarginal)))

ColHPD <- cbind(
  At = 1:(length(ColMergeRes$marginals.fixed)+length(ColRandmode)),
  mode = c(unlist(lapply(ColMergeRes$marginals.fixed, inla.mmarginal)),
           unlist(ColRandmode)),
  low = unlist(c(lapply(ColFixedHPD.l, function(x) x[1,"low"]),
                 lapply(ColRandHPD.l, lapply, function(x) x[1,"low"]))),
  high = unlist(c(lapply(ColFixedHPD.l, function(x) x[1,"high"]),
                 lapply(ColRandHPD.l, lapply, function(x) x[1,"high"])))
)

dimnames(ColHPD)[[1]] <- gsub("\\.", " ", gsub(".LV.*", "", dimnames(ColHPD)[[1]]))
dimnames(ColHPD)[[1]][grep("Pilos", dimnames(ColHPD)[[1]])] <- paste(dimnames(ColHPD)[[1]][grep("Pilos", dimnames(ColHPD)[[1]])], 1:4, sep=" ")
dimnames(ColHPD)[[1]][grep("Poly", dimnames(ColHPD)[[1]])] <- paste(dimnames(ColHPD)[[1]][grep("Poly", dimnames(ColHPD)[[1]])], 1:3, sep=" ")
```

And the plot show femur length has an effect, i.e. ants with long legs prefer feral mammal dung.

```{r PlotColRes}
par(mar=c(4.1,9,1,1))
plot(ColHPD[-1, "mode"], ColHPD[-1, "At"], xlim=range(ColHPD[-1, c("low", "high")]),
     ann=FALSE, axes=FALSE)
segments(ColHPD[-1, "low"], ColHPD[-1, "At"], ColHPD[-1, "high"], ColHPD[-1, "At"])
abline(v=0)
axis(1)
axis(2, dimnames(ColHPD)[[1]][-1], at=ColHPD[-1, "At"], las=1)

```


