---
title: "HierOrdwithCovariates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HierOrdwithCovariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## The Model

This vignette includes covariates into the hierarchical ordination. We have a response $Y_{ij}$ for rows (=sites) $i$ and columns (=species) $j$, where $Y$ is assumed to be from a distribution in the exponential family. Then $E(Y_{ij}) = g^{-1}(\eta_{ij})$. Thus far we have a standard GLM. The fun begins with $\eta$.

We have a latent variable model, so 

$$
\eta_{ij} = \sum_{l=1}^L \gamma_{il} \Sigma_{ll} z_{jl}
$$

with the aim of modelling $\gamma_{il}$ and $z_{jl}$ further. We paramerterise this so that $Var(\gamma_{il}) = Var(z_{jl})=1$, and $\Sigma$ is a diagonal matrix of variances.

Here we will 

For the moment, we are sticking to this model, and also using $L=1$.

## Trying to get this to run.

```{r setup}
library(HierOrd)
library(INLA)
library(coda)
# library(mcmcplots)

```

We will simulate some data from a Poisson distribution.

```{r SimData}
set.seed(42)
SimDat <- SimulateData(NRows = 100, NCols = 50, Sigma = 0.6, NLVs = 1,
                       ColEff = 0.5, RowEff = 1,  Intercept = 4)
str(SimDat)

```

We have a matrix of counts, and the true row and column covariates (which are simulated from a standard normal). We also get the true row and column scores.


The hyperparameters aren't so important at the moment. The model is just for `Roweps.LV1`, with `Wt.Roweps.LV1` as a weight: this is just the current column scores. They will be updated every iteration. `col.formula` has the same format.

Now we run the model

```{r RunMCMC}

Res <-FitHierOrd(Y=SimDat$Counts, rowdata=SimDat$RowCov, 
                 coldata=SimDat$ColCov, nLVs = 1, 
                  family = "Poisson", NBurnin = 10, NIter = 200, NChains = 4, 
#                  family = "Poisson", NBurnin = 1, NIter = 2, NChains = 2, 
                  INLAobj=TRUE)

```


We can check the mixing. Columns first, for 3 rather arbitrary choices.

```{r LookatColMixing, fig.height=8, fig.width=8}
plot(Res$col[,46:48])
```

And rows. Both look OK, although they should probably run for longer.

```{r LookatRowMixing, fig.height=8, fig.width=8}
plot(Res$row[,46:48])
# mcmcplots::caterplot(Res$row, reorder = FALSE)
# mcmcplots::caterplot(Res$col, reorder = FALSE)

```

Now we plot against true scores. The estimates are shrunk towards 0. This might partly be the priors, and also the signs changing.

```{r PotTrueRow, fig.width=8}
RowMean <- summary(Res$row)$statistics[,"Mean"]
RowCI <- summary(Res$row)$quantiles

ColMean <- summary(Res$col)$statistics[,"Mean"]
ColCI <- summary(Res$col)$quantiles

par(mfrow=c(1,2))
plot(SimDat$TrueRowScores, RowMean, ylim=range(RowCI), main="Row")
segments(SimDat$TrueRowScores, RowCI[,"2.5%"], SimDat$TrueRowScores, RowCI[,"97.5%"])
abline(0,1, col=2); abline(0,-1, col=2)

plot(SimDat$TrueColScores, ColMean, ylim=range(ColCI), main="Column")
segments(SimDat$TrueColScores, ColCI[,"2.5%"], SimDat$TrueColScores, ColCI[,"97.5%"])
abline(0,1, col=2); abline(0,-1, col=2)


```

Now the row and column covariates.

```{r RowCovResults, warning=FALSE}

RowMergeRes <- inla.merge(do.call(c, Res$rowinla))
summary(RowMergeRes)$fixed
```


```{r RowCovResults}

ColMergeRes <- inla.merge(do.call(c, Res$colinla))
summary(ColMergeRes)$fixed

ColMergeRes$summary.lincomb.derived
plot(ColMergeRes$summary.lincomb.derived, ColMean)

plot(SimDat$TrueColScores, ColMergeRes$summary.lincomb.derived$mean, ylim=range(ColCI), main="Column")
# segments(SimDat$TrueColScores, ColCI[,"2.5%"], SimDat$TrueColScores, ColCI[,"97.5%"])
abline(0,1, col=2); abline(0,-1, col=2)

```
